Symbolic Trajectories

### 2012-08-17 Algebras/SymbolicTrajectory/st_impl.tex
*** empty log message ***

If the \texttt{matches} operator is called without a condition, no more work is required. Otherwise, for every unit pattern, we have to build another vector of sets containing every cardinality that unit pattern could assume. It is still impossible to perform the condition evaluation or the rewriting, but by combining these cardinality candidates, we acquire the set of all matching sequences (a set of sets of integers). Similar data structures are applied for evaluating conditions and rewriting sequences.

In addition, the number of the final state (integer), the currently active states (set of integers), the current unit label, its id (integer) and the highest unit label id are stored.

Finally, the NFA holds a set of expressions (strings) which have already been evaluated as \ic{true} or \ic{false}, respectively (avoiding multiple evaluation of the same expression in \secondo).

@@

\begin{algorithm}[ht]
  \caption{\label{alg:buildNFA}\ic{buildNFA$(p)$}}
    \KwIn{$p$ - a vector of $n$ Unit Patterns.}
    \KwResult{$\delta$ - the NFA transitions.}
    let $h_0=h_1=h_2=-1$\;
    \For(\tcp*[f]{\scriptsize{loop through pattern}}){$i=0$ \KwTo $n-1$}{
      $\delta_{i,i}\leftarrow i+1$\;
      \uIf{$p_i\notin\{\ast,+,((\dots))\}\ \mathbf{or}\ p_i.ivs\neq\emptyset\ \mathbf{or}\ p_i.lbs\neq\emptyset\ \mathbf{or}\ i=n-1$}{
        \If{$h_0=i-1$ $\mathbf{or}$ $i=n-1$}{
          \For(\tcp*[f]{\scriptsize{step 1}}){$j=h_1+1$ \KwTo $i-1$}{
            $\delta_{j,i}\leftarrow i+1$\;
            \For(\tcp*[f]{\scriptsize{after last match}}){$k=j$ \KwTo $i$}{
              \ic{$\delta_{j,k}\leftarrow j$}\;
              \lFor{$m=j$ \KwTo $i$}{$\delta{j,k}\leftarrow m$\;}
              \If{$p_i=\ast$ $\mathbf{and}$ $i=n-1$}{$\delta_{j,k}\leftarrow n$\;}
            }
          }
          \If(\tcp*[f]{\scriptsize{step 2}}){$h_1>-1$}{
            \lFor{$j=h_1+1$ \KwTo $i$}{$\delta_{h_1,h_1}\leftarrow j$}\;
            \lIf{$p_i=\ast$ $\mathbf{and}$ $i=n-1$}{$\delta_{h_1,h_1}\leftarrow n$\;}
          }
          \If(\tcp*[f]{\scriptsize{step 3}}){$h_2<h_1-1$}{
            \For{$j=h_2+1$ \KwTo $h_1-1$}{
              \lFor{$k=h_1+1$ \KwTo $i$}{$\delta_{j,h_1}\leftarrow k$\;}
              \If{$p_i=\ast$ $\mathbf{and}$ $i=n-1$}{$\delta_{j,h_1}\leftarrow n$\;}
            }
          }
        }
        \lIf{$p_i\in\{+,((\dots))\}$}{$\delta_{i,i}\leftarrow i$\;}
        \ic{$h_2=h_1$}\;
        \ic{$h_1=i$}\;
      }
      \lElseIf{$p_i=\ast$}{$h_0=i$\;}
      \ElseIf{$p_i\in\{+,((\dots))\}$}{
        \ic{$\delta(i,i)\leftarrow i$}\;
        \ic{$h_2=h_1$}\;
        \ic{$h_1=i$}\;
      }
    }
    \lIf{$p_{n-1}\in\{+,\ast\}$}{$\delta_{n-1,n-1}\leftarrow n-1$\;}
\end{algorithm}

@@

\begin{algorithm}[ht]
  \caption{\label{alg:match}\ic{match$(\delta, ml)$}}
    \KwIn{$\delta$ - an NFA transition function\;
      \hspace{1.11cm}$ml$ - an MLabel.}
    \KwOut{the matching result - a boolean.}
    let $S_{cur}=\{0\}$\;
    \ForEach{ULabel $ul\in ml$}{
      let $S_{old}=S_{cur}$\;
      \ForEach{$s\in S_{old}$}{
        \For(\tcp*[f]{\scriptsize{no need to look back}}){$i=s$ \KwTo $n-1$}{
          \If{$\delta_{s,i}\neq\emptyset\ \mathbf{and}\ ul\ \textnormal{matches}\ p_i$}{
            $S_{cur}\leftarrow\delta_{s,i}$\;
            $M_s\leftarrow ulId$\;
          }
        }
      }      
      \lIf{$S_{cur}=\emptyset$}{\Return \ic{false}\;}
    }
    \lIf{$n\in S_{cur}$}{\Return \ic{true}\;}
    \lElse{\Return \ic{false}\;}
\end{algorithm}

Initially, the only active state of the NFA is 0. For each unit label, we perform an update of the currently active states ($S_{cur}$, for brevity). More exactly, the algorithm scans the active states, checks every possible transition from one active state to itself or a higher state, and inserts the new states into $S_{cur}$ if and only if the transition is valid (i.e., $k\in\delta(i,j)$ means that, being in state $i$, state $k$ becomes active if and only if the $j$th unit pattern matches the current unit label).

@@

Therefore, we cannot retrieve useful information from a condition in order to reduce the time consumption. For instance, although it is obvious that trying to match the input \texttt{X $\ast$ Y $\ast$ // X.card = 100} and a moving label with 50 components will fail, the evaluation of the conditions cannot take place until we know all the possible replacements for \texttt{X.card}.

For realizing this, we need to compute a set of cardinality candidates $C_i$ (as hinted in \ref{sec:class condition}) for every unit pattern id, which is done with the help of the matching sets ($M_s$ in algorithm \ref{alg:match}). We distinguish three cases:
\begin{itemize}
  \item a simple pattern: $C_i=\{1\}$
  \item a sequence pattern surrounded by two proper matches: $C_i=M_{i+1}-M_{i-1}$\footnote{For two sets $A, B\subseteq \mathbb{N}_0$, we denote the set of all non-negative differences $\{b-a\colon a\in A, b\in B, a\leq b\}$ by $B-A$}
  \item a sequence pattern besides other sequence patterns: $C_{i}=\{1,\dots,\max\{M_{n}-M_{p}\}\}$, with $n$ and $p$ denoting the next and the previous position of simple patterns, respectively
\end{itemize}

@@

The algorithm performs a combinatorial task, i.e., the cardinality candidates of every unit pattern are combined in every possible way. In line 12, we check the sum of the $n-1$ chosen candidates. If it already exceeds the moving label's size, the current combination is ignored. Otherwise, we consider the non-negative difference $\left|ml\right|-cSum$ which, if it is an element of $C_m$ (i.e., a legal cardinality for the unit pattern $p_m$), is the number that completes the $n-1$ candidates such that the sum of cardinalities equals $\left|ml\right|$. Now, the vector $cards$ contains a valid cardinality sequence, e.g., $cards=(1,0,25,1)$ for a pattern consisting of four units means that the first unit pattern matches the first unit label (0), the second unit pattern (which obviously is a $\ast$) is skipped, the third one matches the sequence of unit labels 1 to 25, and the last one matches the last unit label, 26. Finally, we translate this vector into a sequence indicating the unit label id where the respective matching starts. For the above example, this is $sq=\{0,1,1,26\}$. Every such sequence is stored as a multiset (since $sq_{i-1}\leq sq_i$ holds for $1\leq i<\left|sq\right|$) and inserted into $S$.

\subsubsection{Condition Evaluation}\label{sec:condition evaluation}
Both operators, \texttt{matches} and \texttt{rewrite}, take advantage of $S$. The next step for \texttt{rewrite} is to scan $S$, eliminating the sequences which do not fulfill the set of specified conditions in order to obtain the final set of sequences for rewriting the pattern.

However, we suppose the operator \texttt{matches} is invoked, with the consequence that we need to verify whether there is a sequence in $S$ that complies with every condition. This test is performed in algorithm \ref{alg:conditionsmatch}.

\begin{algorithm}[ht]
  \caption{\label{alg:conditionsmatch}\ic{conditionsMatch$(conds, S, ml)$}}
    \KwIn{$conds$ - a vector of conditions\;
      \hspace{1.11cm}$S$ - as aforementioned\;
      \hspace{1.11cm}$ml$ - an MLabel.}
    \KwOut{the verification result - a boolean.}
    \lIf{$conds=\emptyset$}{\Return \ic{true}\;}
    \For{$i=0$ \KwTo $\left|conds\right|-1$}{
      let \ic{proceed} $=$ \ic{false}\;
      \While{$S\neq\emptyset\ \mathbf{and}\ $proceed $=$ false}{
        \ic{proceed} $=$ \ic{false}\;
        let $sq\in S$\;
        %\ic{buildCondMatchings$(i,sq)$}\;
        \uIf{evaluateCond$(ml,i,sq)=$ false}{
          $S=S\setminus sq$\;
          $i=0$\;
        }
        \lElse{\ic{proceed} $=$ \ic{true}\;}
      }
      \lIf{proceed $=$ false}{\Return \ic{false}\;}
    }
    \Return \ic{true}\;
\end{algorithm}

The outer loop scans the conditions while the inner loop processes the sequences. If a condition evaluation fails (l. 7), the corresponding sequence is deleted, and we have to return to the first condition (l. 9) and to choose another sequence (l. 6). Only if an evaluation is successful, the sequence remains a solution candidate, and we proceed to the next condition. Algorithm \ref{alg:conditionsmatch} returns \ic{false} if and only if there is no element of $S$ which fulfills all conditions.

@@

In order to avoid repeated evaluations of the same term -- which may happen if the pattern contains many stars, or if some labels occur frequently in the moving label -- every analyzed term and its corresponding boolean result are stored into a set. Thus, a substituted condition is only sent to \secondo\ if the aforementioned set does not contain it.\newpage

>>>>

If the \texttt{matches} operator is called without a condition, no more data structure is required. Otherwise, for every unit pattern, we have to build another vector of sets containing every cardinality that unit pattern could assume. It is still impossible to perform the condition evaluation or the rewriting, but by combining these cardinality candidates, we acquire the set of all matching sequences (a set of multisets of integers). Similar data structures are applied for evaluating conditions and rewriting sequences.

In addition, the NFA holds a set of expressions (strings) which have already been evaluated as \ic{true} or \ic{false}, respectively (avoiding multiple evaluation of the same expression in \secondo), together with the corresponding result.

@@

\begin{algorithm}[ht]
  \caption{\label{alg:buildNFA}\ic{buildNFA$(p)$}}
    \KwIn{$p$ - a vector of $n$ Unit Patterns.}
    \KwResult{$\delta$ - the NFA transitions.}
    let $h_0=h_1=h_2=-1$\;
    \For(\tcp*[f]{\scriptsize{loop through pattern}}){$i=0$ \KwTo $n-1$}{
      $\delta_{i,i}\leftarrow i+1$\;
      \uIf{$p_i\notin\{\ast,+,((\dots))\}\ \mathbf{or}\ p_i.ivs\neq\emptyset\ \mathbf{or}\ p_i.lbs\neq\emptyset\ \mathbf{or}\ i=n-1$}{
        \If{$h_0=i-1$ $\mathbf{or}$ $i=n-1$}{
          \For(\tcp*[f]{\scriptsize{step 1}}){$j=h_1+1$ \KwTo $i-1$}{
            $\delta_{j,i}\leftarrow i+1$\;
            \For(\tcp*[f]{\scriptsize{after last match}}){$k=j$ \KwTo $i$}{
              \ic{$\delta_{j,k}\leftarrow j$}\;
              \lFor{$m=j$ \KwTo $i$}{$\delta_{j,k}\leftarrow m$\;}
              \If{$p_i=\ast$ $\mathbf{and}$ $i=n-1$}{$\delta_{j,k}\leftarrow n$\;}
            }
          }
          \If(\tcp*[f]{\scriptsize{step 2}}){$h_1>-1$}{
            \lFor{$j=h_1+1$ \KwTo $i$}{$\delta_{h_1,h_1}\leftarrow j$}\;
            \lIf{$p_i=\ast$ $\mathbf{and}$ $i=n-1$}{$\delta_{h_1,h_1}\leftarrow n$\;}
          }
          \If(\tcp*[f]{\scriptsize{step 3}}){$h_2<h_1-1$}{
            \For{$j=h_2+1$ \KwTo $h_1-1$}{
              \lFor{$k=h_1+1$ \KwTo $i$}{$\delta_{j,h_1}\leftarrow k$\;}
              \If{$p_i=\ast$ $\mathbf{and}$ $i=n-1$}{$\delta_{j,h_1}\leftarrow n$\;}
            }
          }
        }
        \lIf{$p_i\in\{+,((\dots))\}$}{$\delta_{i,i}\leftarrow i$\;}
        \ic{$h_2=h_1$}\;
        \ic{$h_1=i$}\;
      }
      \lElseIf{$p_i=\ast$}{$h_0=i$\;}
      \ElseIf{$p_i\in\{+,((\dots))\}$}{
        \ic{$\delta(i,i)\leftarrow i$}\;
        \ic{$h_2=h_1$}\;
        \ic{$h_1=i$}\;
      }
    }
    \lIf{$p_{n-1}\in\{+,\ast\}$}{$\delta_{n-1,n-1}\leftarrow n-1$\;}
\end{algorithm}

@@

\begin{algorithm}[ht]
  \caption{\label{alg:match}\ic{match$(\delta, ml)$}}
    \KwIn{$\delta$ - an NFA transition function\;
      \hspace{1.11cm}$ml$ - an MLabel.}
    \KwOut{the matching result - a boolean.}
    let $S_{cur}=\{0\}$\;
    \ForEach{ULabel $ul\in ml$}{
      let $S_{old}=S_{cur}$\;
      \ForEach{$s\in S_{old}$}{
        \For(\tcp*[f]{\scriptsize{no need to look back}}){$i=s$ \KwTo $n-1$}{
          \If{$\delta_{s,i}\neq\emptyset\ \mathbf{and}\ ul\ \textnormal{matches}\ p_i$}{
            $S_{cur}\leftarrow\delta_{s,i}$\;
            $M_s\leftarrow ulId$\;
          }
        }
      }      
      \lIf{$S_{cur}=\emptyset$}{\Return \ic{false}\;}
    }
    \lIf(\tcp*[f]{\scriptsize{final state active?}}){$n\in S_{cur}$}{\Return \ic{true}}
    \lElse{\Return \ic{false}\;}
\end{algorithm}

Initially, the only active state of the NFA is 0. For each unit label, we perform an update on the set of currently active states ($S_{cur}$, for brevity). More exactly, the algorithm scans the active states, checks every possible transition from one active state to itself or a higher state, and inserts the new states into $S_{cur}$ if and only if the transition is valid (i.e., $k\in\delta(i,j)$ means that, being in state $i$, state $k$ becomes active if and only if the $j$th unit pattern matches the current unit label).

@@

Therefore, we are not able to retrieve useful information from a condition in order to reduce the time consumption. For instance, although it is obvious that trying to match the input \texttt{X $\ast$ Y $\ast$ // X.card = 100} and a moving label with 50 components will fail, the evaluation of the conditions cannot take place until we know the appropiate value candidates to put in for \texttt{X.card}.

For realizing this, we need to compute a set of cardinality candidates $C_i$ for every unit pattern id $i$, which is done with the help of the matching sets ($M_s$ in algorithm \ref{alg:match}). We distinguish three cases:
\begin{itemize}
  \item a simple pattern: $C_i=\{1\}$
  \item a sequence pattern surrounded by two proper matches: $C_i=M_{i+1}-M_{i-1}$\footnote{For two sets $A, B\subseteq \mathbb{N}_0$, we denote the set of all non-negative differences $\{b-a\colon a\in A, b\in B, a\leq b\}$ by $B-A$}
  \item a sequence pattern besides other sequence patterns: $C_{i}=\{1,\dots,\max\{M_{n}-M_{p}\}\}$, with $n$ and $p$ denoting the next and the previous position of simple patterns, respectively
\end{itemize}

@@

The algorithm performs a combinatorial task, i.e., the cardinality candidates of every unit pattern are combined in every possible way. In line 12, we check the sum of the $n-1$ chosen candidates. If it already exceeds the moving label's size, the current combination is ignored. Otherwise, we consider the non-negative difference $\left|ml\right|-cSum$ which, if it is an element of $C_m$ (i.e., a valid cardinality for the unit pattern $p_m$), is the number that completes the $n-1$ candidates such that the sum of cardinalities equals $\left|ml\right|$. Now, the vector $cards$ contains a valid cardinality sequence, e.g., $cards=(1,0,25,1)$ for a pattern consisting of four units means that the first unit pattern matches the first unit label (0), the second unit pattern (which obviously is a $\ast$) is skipped, the third one matches the sequence of unit labels 1 to 25, and the last one matches the last unit label, 26. Finally, we translate this vector into a sequence indicating the unit label id where the respective matching starts. For the above example, this is $sq=\{0,1,1,26\}$. Every such sequence is stored as a multiset (since $sq_{i-1}\leq sq_i$ holds for $1\leq i<\left|sq\right|$) and inserted into $S$.

Since algorithm \ref{alg:buildsequences} is heavily time-consuming, this version is only executed if the operator \texttt{rewrite} is called. The other case is discussed in \ref{sec:condition evaluation}.

\subsubsection{Condition Evaluation}\label{sec:condition evaluation}
The next step for \texttt{rewrite} is to scan $S$, eliminating the sequences which do not fulfill the set of specified conditions in order to obtain the final set of sequences for rewriting the pattern.

However, we suppose the operator \texttt{matches} is invoked, with the consequence that it suffices to verify whether there is a sequence that complies with every condition, which is performed in algorithm \ref{alg:conditionsmatch}. Hence, it is not necessary to compute $S$ as seen in algorithm \ref{alg:buildsequences}, and with slight modifications we can use the latter to create only one sequence at a time.

\begin{algorithm}[ht]
  \caption{\label{alg:conditionsmatch}\ic{conditionsMatch$(conds, S, ml)$}}
    \KwIn{$conds$ - a vector of conditions\;
      \hspace{1.11cm}$ml$ - an MLabel.}
    \KwOut{the verification result - a boolean.}
    \lIf{$conds=\emptyset$}{\Return \ic{true}\;}
    let $sq=$ \ic{getNextSeq$()$}\;
    let \ic{proceed} $=$ \ic{false}\;
    \For{$i=0$ \KwTo $\left|conds\right|-1$}{
      \While{$sq\neq\emptyset\ \mathbf{and}\ $proceed $=$ false}{
        \ic{proceed} $=$ \ic{false}\;
        \uIf{evaluateCond$(ml,i,sq)=$ false}{
          $sq=$ \ic{getNextSeq$()$}\;
          $i=0$\;
        }
        \lElse{\ic{proceed} $=$ \ic{true}\;}
      }
      \lIf{proceed $=$ false}{\Return \ic{false}\;}
    }
    \Return \ic{true}\;
\end{algorithm}

The outer loop scans the conditions while the inner loop processes the sequences. If a condition evaluation fails (l. 7), the next sequence is requested, and we have to return to the first condition (l. 9) and to request the next sequence (l. 8; this function either returns the next sequence, or, if no more sequences are available, an empty set). Only if an evaluation is successful, the sequence remains a solution candidate, and we proceed to the next condition. Algorithm \ref{alg:conditionsmatch} returns \ic{false} if and only if there is no sequence returned by \ic{getNextSeq$()$} which fulfills all conditions.

@@

In order to avoid repeated evaluations of the same term -- which may happen if the pattern contains many stars, or if some labels occur repeatedly in the moving label -- every analyzed term and its corresponding boolean result are stored into a set. Thus, a substituted condition is only sent to \secondo\ if the aforementioned set does not contain it.

\subsubsection{Filtering the Sequences}
At this point, the operator \texttt{matches} has finished its task, in contrast to \texttt{rewrite}. As mentioned before, in the latter we have computed the set $S$ of all matching sequence candidates. Unlike in the previous subsection, it is not enough to decide whether there is a sequence matching every condition, but we need to find all the sequences that match every condition -- that is, to filter the set of sequences.

The filtering works similarly to algorithm \ref{alg:conditionsmatch}, i.e., for every sequence in $S$, we check whether it fulfills every condition. If it does, it stays in $S$, otherwise it is deleted. Now, all information for rewriting the pattern is available.
\newpage
